### 2020-02-14 thesis/content/chapter-related-work.tex
minor additions to wl section
currently still working on WL illustrations...

The class of \ac{lta} problems was first described by \citet{Melnikov2016}.
There an input instance is understood as a composition $\bm{c}$ of so-called constituents $c_i \in \bm{c}$, i.e.\ as a variable-size multiset with $n = |\bm{c}|$.
The assumption in \ac{lta} problems is that for all constituents $c_i$ a local score $y_i \in \mathcal{Y}$ is either given or computable.
The set of those local scores should be indicative of the overall score $y \in \mathcal{Y}$ of the composition $\bm{c}$.
\ac{lta} problems typically require two subproblems to be solved:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Aggregation:}
		A variadic aggregation function $\mathcal{A}: \mathcal{Y}^{*} \to \mathcal{Y}$ that estimates composite scores has to be learned, i.e.\@ $y_i \approx \hat{y} = \mathcal{A}(y_{1}, \dots, y_{n})$.
		Typically the aggregation function $\mathcal{A}$ should be associative and commutative to fit with the multiset-structure of compositions.
	\item \textbf{Disaggregation:}
		In case the constituent scores $y_i$ are not given, they have to be derived from a constituent representation, e.g.\ a vector $x_i \in \mathcal{X}$.
		To learn this derivation function $f: \mathcal{X} \to \mathcal{Y}$, only the constituent vectors ${\{x_i\}}_{i = 1}^{n}$ and the composite score $y$ is given.
		Thus the constituent scores $y_i$ need to be \textit{disaggregated} from $y$ in order to learn $f$.
\end{enumerate}
\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{gfx/related-work/lta-overview.pdf}
	\caption{Overview of the structure of LTA for multiset compositions.}\label{fig:related:lta-overview}
\end{figure}
Overall \ac{lta} can be understood as the joint problem of learning the aggregation function $\mathcal{A}$ and the local score derivation function $f$.
Two main approaches to represent the aggregation function in \ac{lta} problems have been explored.

@@

Since t-norms and t-conorms are commutative and associative they can also be applied to non-empty sets of arbitrary size, i.e. $T(\{ y_1, \dots, y_n \}) = T(y_1, T(\{ y_2, \dots, y_n \}))$ with fixpoint $T(\{ y \}) = y$.
Using this extension, a uninorm $U_\lambda$ can be applied to sets which turns it into a parameterized aggregation function $\mathcal{A}_\lambda: {[0, 1]}^* \to [0, 1]$.
In this simple model the \ac{lta} aggregation problem boils down to the optimization of $\lambda$.
The \ac{lta} disaggregation problem is solved by jointly optimizing a \acf{lm}, i.e.\ the constituent scores ${\{ y_i \in [0, 1] \}}_{c_i \in \bm{c}}$ are described by $y_i = {\left( 1 + \exp(-\theta^\top x_i) \right)}^{-1}$.
Overall an \ac{lta} model is therefore described by the uninorm parameter $\lambda$ and the regression coefficients $\theta$.

@@

To deal with varying composition sizes $n$, the weights ${\{ \lambda_i \}}_{i = 1}^{n}$ can however not be statically assigned.
Instead they are interpolated using a so-called \ac{bum} function $q: [0, 1] \to [0, 1]$.
It takes constituent positions that are normalized to the unit interval, i.e.\ $\frac{i}{n} \in [0, 1]$.
The \ac{bum} function $q$ is then used to interpolate a weight for any normalized sort position via $\lambda_i := q\left(\frac{i}{n}\right) - q\left(\frac{i - 1}{n}\right)$.
Because $q$ is monotone with $q(0) = 0$ and $q(1) = 1$, it always holds that ${\|\lambda\|}_1 = q(1) - q(0) = 1$. % chktex 21
Using this model, the aggregation problem boils down to optimizing the shape of $q$.

@@

\subsubsection{The 1-dimensional \acs{wl} algorithm}
In the 1-dimensional \ac{wl} algorithm, a color is assigned to each vertex of a graph.
If the vertices $\mathcal{V}_G$ of the input graph $G$ are already labeled via some labeling $l_G: \mathcal{V}_G \to L$, those labels can be used as the initial graph coloring $\chi_{G,1}^0 = l_G$.
For unlabeled graphs a constant coloring is used instead, i.e.\ $\forall v \in \mathcal{V}_G: \chi_{G,1}^0(v) = \texttt{A}$ for some initial color $\texttt{A} \in \mathcal{C}$.
In each iteration of the 1-\acs{wl} color refinement algorithm, the following neighborhood aggregation scheme is used to compute a new color for all vertices:
\begin{align}
	\chi_{G,1}^{i+1}(v) := h\left(\chi_{G,1}^{i}(v), \{ \chi_{G,1}^{i}(u) | u \in \Gamma_{G,1}(v) \}\right)
\end{align}

>>>>

The class of \ac{lta} problems was first described by \citet{Melnikov2016}.
There an input instance is understood as a composition $\bm{c} = \ldblbrace c_1, \dots, c_n \rdblbrace$ of so-called constituents, i.e.\ as a variable-size multiset (denoted as $\ldblbrace \cdot \rdblbrace$).
The assumption in \ac{lta} problems is that for all constituents $c_i \in \bm{c}$ a local score $y_i \in \mathcal{Y}$ is either given or computable.
The set of those local scores should be indicative of the overall score $y \in \mathcal{Y}$ of the composition $\bm{c}$.
\ac{lta} problems typically require two subproblems to be solved:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Aggregation:}
		A variadic aggregation function $\mathcal{A}: \mathcal{Y}^{*} \to \mathcal{Y}$ that estimates composite scores has to be learned, i.e.\@ $y_i \approx \hat{y} = \mathcal{A}(y_{1}, \dots, y_{n})$.
		Typically the aggregation function $\mathcal{A}$ should be associative and commutative to fit with the multiset-structure of compositions.
	\item \textbf{Disaggregation:}
		In case the constituent scores $y_i$ are not given, they have to be derived from a constituent representation, e.g.\ a vector $x_i \in \mathcal{X}$.
		To learn this derivation function $f: \mathcal{X} \to \mathcal{Y}$, only the constituent vectors ${\ldblbrace x_i \rdblbrace}_{i = 1}^{n}$ and the composite score $y$ is given.
		Thus the constituent scores $y_i$ need to be \textit{disaggregated} from $y$ in order to learn $f$.
\end{enumerate}
\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{gfx/related-work/lta-overview.pdf}
	\caption{Overview of the structure of LTA for multiset compositions.}\label{fig:related:lta-overview}
\end{figure}
Overall \ac{lta} can be understood as the joint problem of learning the aggregation function $\mathcal{A}$ and the local score derivation function $f$.
Two main approaches to represent the aggregation function in \ac{lta} problems have been explored.

@@

Since t-norms and t-conorms are commutative and associative they can also be applied to non-empty sets of arbitrary size, i.e. $T(\ldblbrace y_1, \dots, y_n \rdblbrace) = T(y_1, T(\ldblbrace y_2, \dots, y_n \rdblbrace))$ with fixpoint $T(\ldblbrace y \rdblbrace) = y$.
Using this extension, a uninorm $U_\lambda$ can be applied to sets which turns it into a parameterized aggregation function $\mathcal{A}_\lambda: {[0, 1]}^* \to [0, 1]$.
In this simple model the \ac{lta} aggregation problem boils down to the optimization of $\lambda$.
The \ac{lta} disaggregation problem is solved by jointly optimizing a \acf{lm}, i.e.\ the constituent scores ${\ldblbrace y_i \in [0, 1] \rdblbrace}_{c_i \in \bm{c}}$ are described by $y_i = {\left( 1 + \exp(-\theta^\top x_i) \right)}^{-1}$.
Overall an \ac{lta} model is therefore described by the uninorm parameter $\lambda$ and the regression coefficients $\theta$.

@@

To deal with varying composition sizes $n$, the weights $\lambda_1, \dots, \lambda_n$ can however not be statically assigned.
Instead they are interpolated using a so-called \ac{bum} function $q: [0, 1] \to [0, 1]$.
It takes constituent positions that are normalized to the unit interval, i.e.\ $\frac{i}{n} \in [0, 1]$.
The \ac{bum} function $q$ is then used to interpolate a weight for any normalized sort position via $\lambda_i := q\left(\frac{i}{n}\right) - q\left(\frac{i - 1}{n}\right)$.
Because $q$ is monotone with $q(0) = 0$ and $q(1) = 1$, it always holds that ${\|\lambda\|}_1 = q(1) - q(0) = 1$. % chktex 21
Using this model, the aggregation problem boils down to optimizing the shape of $q$.

@@

\subsubsection{The 1-dimensional \acs{wl} algorithm}
In the 1-dimensional \ac{wl} algorithm, a color is assigned to each vertex of a graph.
If the vertices $\mathcal{V}_G$ of the input graph $G$ are already labeled via some labeling $l_G: \mathcal{V}_G \to L$, those labels can be used as the initial graph coloring $\chi_{G,1}^0 = l_G$.
For unlabeled graphs a constant coloring is used instead, i.e.\ $\forall v \in \mathcal{V}_G: \chi_{G,1}^0(v) = \texttt{A}$ for some initial color $\texttt{A} \in \mathcal{C}$.
In each iteration of the 1-\acs{wl} color refinement algorithm, the following neighborhood aggregation scheme is used to compute a new color for all vertices:
\begin{align}
	\chi_{G,1}^{i+1}(v) := h\left(\chi_{G,1}^{i}(v), \ldblbrace \chi_{G,1}^{i}(u)\, |\, u \in \Gamma_{G}(v) \rdblbrace\right) \text{,}
\end{align}
with $\Gamma_G(v)$ denoting the set of adjecent vertices of $v \in \mathcal{V}_G$ and $h: \mathcal{C}^* \to \mathcal{C}$ denoting a perfect hash function that assigns a unique color to each finite combination of colors.
In practice $h$ is usually defined lazily by using $\mathcal{C} = \mathbb{N}$ and non-deterministically enumerating color combinations in the order in which they are hashed s.t.\ a new color is introduced every time a previously unseen color combination is queried.
